---
title: 'Data transformation 1'
subtitle: 'Introduction to RStudio'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.3/MathJax.js?config=TeX-MML-AM_CHTML"
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

# Data wrangling

Much of the data wrangling I do daily (and it is daily!) can be done using only a handful of functions, the majority taken from the package `dplyr`, with a few extras from other packages such as `tidyr`.

Essentially, the `dplyr` package provides a bunch of "verbs" (functions) for transforming data. It is similar to SQL if you have done any database work.

It is part of the `tidyverse` set of packages: https://www.tidyverse.org

```{r, echo=FALSE, message=FALSE}
hrc <- read_csv(here::here("data/horizons_river_quality/horizons_river_ecoli.csv"))
```

---

class: middle

# `dplyr`

---

## `dplyr`

Just like `ggplot2` is a grammar of graphics, `dplyr` is a grammar of data manipulation.

It consists of a set of verbs that help us solve the most common data manipulation challenges:

- `select` to choose variables (columns) by name.
- `filter` to choose observations (rows) by value.
- `mutate` to add new variables based of existing variables.
- `summarise` to reduce multiple values down to a single summary.
- `arrange` changes the order of rows.

All of these combine with `group_by` to do all the above operation on subgroups of the data.

---

## `dplyr`

One of the nice things about `dplyr` is it abstracts the language of data processing across the implementation of where the data exist.

It works with `data.frame`'s loaded into memory, as well as remote databases (via SQL, or Apache Spark) using the same set of R commands.

The other nice thing is the consistency:
 - All verbs take in as their first argument a `data.frame`.
 - Then all other arguments describe what to do with the `data.frame`.
 - The result is then a new `data.frame`.

```{r eval=FALSE}
new_data = do_something(old_data, other, stuff)
```
 
---

## Filtering rows with `filter`
 
Let's look at the horizons council river data:
 
```{r}
hrc
```

---

## Filtering rows with `filter`
 
What if we just wanted the rows from 2017 onwards?
 
```{r}
filter(hrc, Date >= '2017-01-01')
```

---

## Filtering rows with `filter`
 
What if we wanted rows with large counts?

```{r}
filter(hrc, Value > 100000)
```

---

## Filtering rows with `filter`
 
What if we wanted really low counts before 2010?

```{r}
filter(hrc, Value < 100, Date <= '2010-12-31')
```

---

class: inverse

## Try yourself

Open up `dplyr01.R` which contains some school roll data.

Use `filter` to find all rows corresponding to Queen Elizabeth College students in Year 9.

---

## Filtering rows with `filter`
 
What if we wanted the extreme low or high counts?

```{r}
filter(hrc, Value < 3 | Value > 100000)
```

---

## Logical expressions in R

The `filter` argument is a logical expression. i.e. it's something that evaluates to `TRUE` or `FALSE`.

For convenience, filter can take more than one logical expression, separated by a comma. In this case they're AND'd together (i.e. rows must satisfy both criteria), but anything within each expression is done first.

If we want to use OR rather than AND, we use the vertical bar `|` character.

i.e. `filter(hrc, Value < 3 | Value > 100000, SiteID == "00081")`

will return rows where the value is very small or very large from Site 00081.

---

## Logical expressions in R

Notice that for 'is equal to' we've been using `==`. The single `=` is for assignment of parameters instead. `dplyr` will warn you about this:

```{r, error=TRUE}
filter(hrc, Value = 4)
```

**This is rare** - usually error messages are rather more opaque!

Reading carefully, and searching the error on google can sometimes help.

---

## Logical expressions in R

Other operations are `!=` (not equal to) and `%in%` (in). e.g.

```{r}
filter(hrc, Value %in% c(3,5))
```

---

class: inverse

## Try yourself

In `dplyr01.R`, add some code to:

1. Find all rows where there are more than 100 Year 9 Asian students.
2. Find all rows where the school name starts with the letter 'G' (hint: you can use School > 'G' for this, but using `str_starts` might be better!)

---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing? We might try:

```{r}
filter(hrc, Symbol != NA)
```

This seems like it might have worked?? It says there are no rows returned...

--

```{r}
filter(hrc, Symbol == NA)
```

Hmm, that doesn't seem right??

---

## Dealing with missing values

.left-code[
The key thing to remember is if a value is missing, then comparing it is senseless:

```{r}
2 == NA
NA == NA
```

Is 2 equal to something missing?

Maybe! We don't know! So the answer is missing.
]

--
.right-plot[
We use `is.na()` instead for this.

```{r}
is.na(2)
is.na(NA)
```
]

RStudio will remind you about this with a yellow warning in the margin, as it's a common problem.
---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing?

```{r}
filter(hrc, !is.na(Symbol))
```

Ok, it seems like we have a few (233) numbers where the count actually means less than or greater than the value given, rather than equal to the value given.

---

class: inverse

## Try yourself

The school roll dataset has no explicit missing values: all rows are complete.

However, it has **implicit** missing values, in that not all level, gender, ethnicity combinations are present at each school. i.e. anything with a 0 is not represented by a row.

Use `filter` to confirm that there are no 0 counts for `Students`.

---

## Arranging rows with `arrange`

Rather than pick rows out, we can sort the data instead.

```{r}
arrange(hrc, Value)
```

---

## Arranging rows with `arrange`

Selecting more than one variable to arrange on will first arrange on the first, then use later variables for ties:

```{r}
arrange(hrc, Value, Date)
```

---

## Arranging rows with `arrange`

To arrange descending rather than ascending, we can use `desc`

```{r}
arrange(hrc, Value, desc(Date))
```

Interestingly, no value of 1 after 2012... This suggests a change in the way things were measured.

---

class: inverse

## Try yourself

Arrange the `roll` rows by number of students, ascending then descending.

---

## Combining `filter` and `arrange`

Let's look at post-2013 values:

```{r}
after2012 = filter(hrc, Date >= '2013-01-01')
arrange(after2012, Value)
```

It seems some of the values recorded as 4 actually mean "less than 4".

---

## Combining `filter` and `arrange`

```{r}
large_values = filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

---

class: inverse:

## Try yourself

1. Find all rows where the ethnicity is Pacific, arranging them by decreasing number of students. (Hint: First filter down to Pacific students, then arrange)

2. Which school has the highest number of International fee paying students at Year 13? (Hint: 'International fee paying' is an option for `EthnicGroup`)

---

## The pipe `%>%`

Each of the verbs in `dplyr` take a dataset in as the first parameter, and return a data frame out again.

Often, an analyses will consist of a bunch of verbs called one after the other, with each one feeding into the next.

```{r, eval=FALSE}
large_values = filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

The temporary dataset `large_values` is really only used here to feed to the `arrange()` verb. We don't really need it.

We could do:

```{r, eval=FALSE}
arrange(filter(hrc, Value > 50000), Site, Date)
```

But you have to read that 'inside out'!

---

class: middle

# The pipe `%>%`

---

## The pipe `%>%`

The pipe operator, `%>%` is specifically designed to improve this. It takes the item on the left of the pipe and places it
as the first argument of the function on the right.

```{r, eval=FALSE}
hrc %>% filter(Value > 5000)
```

Is equivalent to:

```{r, eval=FALSE}
filter(hrc, Value > 5000)
```

But, the pipe version is readable as "Take the dataset `hrc` and filter it so that rows with Value > 5000 remain"

To enter a pipe in RStudio, you can use Ctrl-Shift-M (Command-Shift-M on a Mac).

---

## The pipe `%>%`

This extends naturally:

```{r, eval=FALSE}
hrc %>%
  filter(Value > 5000) %>%
  arrange(Site, Date)
```

reads "Take the dataset `hrc`, filter it so that rows with Value > 5000 remain, then arrange by Site and Date"

When constructing pipelines like this, make sure the `%>%` symbol is at the end of intermediate lines.

.left-code[
Use this:

```{r, eval=FALSE}
hrc %>%
  filter(Value > 5000) %>%
  arrange(Site, Date)
```
]

.right-plot[
Not this:

```{r, eval=FALSE}
hrc
 %>% filter(Value > 5000)
 %>% arrange(Site, Date)
```
]

---

class: inverse

## Try yourself

Convert the previous few exercises to use the pipe.

All commands should start with `roll %>%`

---

## Selecting and renaming columns

Just like we can select rows with `filter`, we can choose columns with `select`.

This is useful for getting rid of columns we don't need, for rearranging columns, and for changing names to
something more convenient.

```{r}
hrc %>% select(Site, Date, Value)
```

---

## Selecting columns with `select`

We can select a range of columns with a colon `:`

```{r}
hrc %>% select(Council, Site:RawValue)
```

---

## Selecting columns with `select`

We can select by column number instead of column name if you like:

```{r}
hrc %>% select(Council, 3:6)
```

Generally it's better to use names though, as the order of columns might be different to what you expect.

---

## Selecting columns with `select`

The `everything()` helper function is useful if you want one column first, then everything else:

```{r}
hrc %>% select(Symbol, Value, everything())
```

---

## Selecting columns with `select`

The `starts_with()` helper is useful if you have a bunch of columns that start similarly:

```{r}
hrc %>% select(Date, Value, starts_with("Site"))
```

---

## Selecting columns with `select`

You can rename a column just by changing it's name in the `select`:

```{r}
hrc %>% select(Site, Date, EColi = Value)
```

---

## Dropping columns

To drop a column use the `-` in front of the name:

```{r}
hrc %>% select(-parameter, -RawValue, -Symbol)
```

---

class: inverse

## Try yourself

1. Rearrange the `roll` data set so that `Students` and `Level` are first.
2. Rename the `School` column to `Name` keeping all other columns.
3. Try part 2 using `rename` instead of `select`.

---

## Renaming columns with `rename`

If you want to rename a column but don't want to specify all the other columns, then `rename()` does this:

```{r}
hrc %>% rename(Name = Site)
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r}
hrc %>% select(SiteID, Date, Value) %>%
  mutate(Log10Value = log10(Value))
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r, message=FALSE}
library(lubridate)
hrc %>% select(SiteID, Date, Value) %>%
  mutate(Year = year(Date), Month = month(Date, label=TRUE))
```

---

.left-code[
## Seasonality?
```{r, label='hrc_season', eval=FALSE}
hrc %>%
  mutate(
    Month = month(Date, label=TRUE)
    ) %>%
  ggplot() +
  geom_boxplot(
    aes(
      x = Month,
      y = Value)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  )
```

We can pipe the result of wrangling into
`ggplot()` as it takes a data.frame as first
argument.

Remember to switch from `%>%` to `+` !!
]

.right-plot[
```{r, ref.label='hrc_season', echo=FALSE, eval=TRUE}
```
]

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
library(lubridate)
hrc %>% select(SiteID, Date, Value) %>%
  mutate(SiteID = as.numeric(SiteID))
```

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
hrc %>% select(SiteID, Site, Date, Value) %>%
  mutate(SiteID = as.numeric(SiteID),
         Site = str_to_title(Site))
```

---

class: inverse

### Try yourself

Open up `dplyr02.R`. You'll see it is creating a new, "clean" dataset using `mutate`.

1. Using the `clean` dataset, create a new column combining ethnicgroup and gender with `paste`.
2. Redo part 1, so that the original two columns are removed.
3. Create a new variable "Age" with the typical age of students in each year level (Year 1 students are typically 5 years old).

---

## How I extracted value, symbol

.left-code-wide[
```{r, label="hrc_extract", eval=FALSE}
hrc %>% select(SiteID, RawValue) %>%
  mutate(
    Symbol = str_extract(RawValue, '[<>]'),
    Value = str_extract(RawValue, '[0-9]+'),
    Value = as.numeric(Value)
    ) %>%
  arrange(Symbol)
```

The patterns `[<>]`, `[0-9]+` are **regular expressions**.

They're a way of specifying how to match parts of text.

Knowing regular expressions really helps with data wrangling - I often google how to do it each time though!
]

.right-out-narrow[
```{r, ref.label="hrc_extract", echo=FALSE, eval=TRUE}
```
]

---

class: inverse

## More exercises

TODO: Add extra stuff here
