---
title: 'Visualisation 1'
subtitle: 'Introduction to RStudio'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
penguins = na.omit(penguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

## Visualisation in R/RStudio

There are three main charting 'systems' in R/RStudio:

- base graphics. Included with the R `base` package. Commands like `plot`, `barplot`, `boxplot`.

- lattice graphics. Included with the R `lattice` package. Commands like `xyplot` etc.

- `ggplot2` graphics. From the `ggplot2` package. **This is what we'll use.**

All have advantages and disadvantages. We teach `ggplot2` as it is the most flexible and consistent
way of achieving publication ready graphics.

---

class: middle

# `ggplot2`

---

## What is `ggplot2`

- `ggplot2` is a package for R by [Hadley Wickham](http://hadley.nz/)

- Based on the **grammar of graphics**

- Uses a consistent 'language' to produce different plots

- Very flexible, particularly for complex plots

- Everything is editable

- **BUT: Quite a lot to type!**

---

.left-code[
## Quite a lot to type
```{r quakes, eval=FALSE}
ggplot(data = quakes) +
  geom_point(
    mapping = aes(
      x = long,
      y = lat
    )
  )
```
]

.right-plot[
```{r ref.label="quakes", echo=FALSE}
```
]

---

## Helpful resources

- The `ggplot2` cheatsheet in RStudio: https://rstudio.org/links/data_visualization_cheat_sheet

- The `ggplot2` book: https://ggplot2-book.org/

- The `ggplot2` page on tidyverse.org: https://ggplot2.tidyverse.org/

- Search stackoverflow.com: https://stackoverflow.com/questions/tagged/ggplot2?sort=faq

---

## What is the grammar of graphics?

Basic idea: Specify plot building blocks and combine them to produce a plot.

- Data. Always use a `data.frame`

- Aesthetics. Mapping of your data to a plot aesthetic (position/colour/fill/shape)

- Geometries. Points/Lines/Boxplots/Histograms etc

- Statistical transformations

- Scales and coordinate systems

- Position adjustments

- **Faceting** (small multiple plots)

---

## Example: Penguins!

Data today were collected by [Dr Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antartica Long Term Ecological Research Network](https://pal.lternet.edu/).

You can get it for R at https://github.com/allisonhorst/palmerpenguins

```{r}
penguins
```

---

.left-code[
```{r peng1, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

.right-plot[
```{r ref.label="peng1", echo=FALSE}
```
]
---

## Recipe

To create a chart we need to:

- Tell R we want a plot with `ggplot()`.

- Tell `ggplot` what data we are plotting with `data=penguins`.

- Choose what type of chart we want by adding a layer (geometry), e.g. `geom_point`.

- Tell the geometry how to map from visual content/features to data with `mapping = aes()`.

All our charts will follow the same basic recipe. We might just add a few more ingredients!

---

.left-code[
## Recipe
```{r peng2, eval=FALSE}
ggplot(data = penguins) + #<<
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
Tell `ggplot` what data we want to plot.
]

.right-plot[
```{r ref.label="peng2", echo=FALSE}
```
]

---

.left-code[
## Recipe
```{r peng3, eval=FALSE}
ggplot(data = penguins) +
  geom_point( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) #<<
```
Choose the type of chart we want.
]

.right-plot[
```{r ref.label="peng3", echo=FALSE}
```
]

---

.left-code[
## Recipe
```{r peng4, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    )
```

Map plot features to data features.
]

.right-plot[
```{r ref.label="peng4", echo=FALSE}
```
]

---

class: inverse

## Try yourself

- Go to RStudio and open `ggplot01.R` in the `01_visualisation` folder.

- If prompted, install packages.

- Run the code to get the plot.

---

.left-code[
## Mapping colour to data
```{r peng7, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = species #<<
      )
    )
```

- Automatically adds a guide or legend.

- `x` and `y` have guides too (the axes).
]

.right-plot[
```{r, ref.label="peng7", echo=FALSE}
```
]

---

class: inverse

## Try yourself

Alter the code in `ggplot01.R` to colour by species.

---

## Remember the Recipe!

- Tell R we want a plot with `ggplot()`.

- Tell `ggplot` what data we are plotting with `data=penguins`.

- Choose what type of chart we want by adding a layer (geometry), e.g. `geom_point()`.

- Tell the geometry how to map from visual content/features to data with `mapping = aes()`.

---

.left-code[
## Remember the Recipe!
```{r peng2b, eval=FALSE}
ggplot(data = penguins) + #<<
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
Tell `ggplot` what data we want to plot.
]

.right-plot[
```{r ref.label="peng2b", echo=FALSE}
```
]

---

.left-code[
## Remember the Recipe!
```{r peng3b, eval=FALSE}
ggplot(data = penguins) +
  geom_point( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) #<<
```
Choose the type of chart we want.
]

.right-plot[
```{r ref.label="peng3b", echo=FALSE}
```
]

---

.left-code[
## Remember the Recipe!
```{r peng4b, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    )
```

Map plot features to data features.
]

.right-plot[
```{r ref.label="peng4b", echo=FALSE}
```
]

---

class: inverse

## Code along!

Follow along as we break down what the code in `ggplot01.R` is doing.

---

class: middle

# Aesthetics

## Setting or mapping features of the plot

---

## Aesthetics

The visual content or features of the plot are **aesthetics**. Things like:

- position of points (`x`, `y`)

- height of bars (`y`)

- colour of points (`colour`, `alpha`)

- colour inside regions (`fill`)

- point type and size (`shape`, `size`)

All of these can either be **set to a single value**, or allowed to **change by mapping them to data**.

---

.left-code[
## Setting a colour
```{r peng5, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      ),
    colour = 'purple' #<<
    )
```

To **set** to a single value, place inside the `geom_*()` function.
]

.right-plot[
```{r, ref.label="peng5", echo=FALSE}
```
]

---

.left-code[
## Setting versus Mapping
```{r peng6, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = 'purple' #<<
      )
    )
```
To **set** to a single value, place inside the `geom_*()` function.

But not inside the `aes()` function!

This is like adding a new column, and setting every entry to `purple`, so you have a grouping variable with one group `purple`.
]

.right-plot[
```{r, ref.label="peng6", echo=FALSE}
```
]

---

.left-code[
## Mapping colour to data
```{r peng7b, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = species #<<
      )
    )
```

To **map to a column of data** place inside the `aes()` function.
]

.right-plot[
```{r, ref.label="peng7b", echo=FALSE}
```
]

---

class: inverse

## Try yourself

Load up `ggplot02.R`.

This code is trying to change the shape of the points to squares, but
we've made a mistake.

Find the mistake and fix it.

---

## Mapping aesthetics

Aesthetics are mapped to data via the `aes()` function passed to the `mapping` parameter of each geometry layer.

What this does is define a map between features of the plot (`x`, `y`, `colour`) and features of our data (columns).

When this is done, `ggplot` automatically adds a scale or guide so you know which styling maps
to which parts of your data.

e.g. the `x` and `y` aesthetics automatically have an axis scale. The `colour` aesthetic will give a legend or colour scale.

**Any time you want a feature of the plot to change based on data, it needs to be in `aes()`**

---

## Different species

```{r}
aes(x = flipper_length_mm, 
    y = body_mass_g,
    colour = species)
```

---

.left-code[
## Different species
```{r peng8, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = species #<<
      )
    )
```
`colour` changes with data (mapping).
]

.right-plot[
```{r, ref.label="peng8", echo=FALSE}
```
]

---

.left-code[
## Different species
```{r peng9, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      size = species #<<
      ),
    alpha = 0.5 #<<
    )
```
`size` changes with data (mapping).

`alpha` is fixed for all points (setting).
]

.right-plot[
```{r, ref.label="peng9", echo=FALSE}
```
]

---

.left-code[
## Different species
```{r peng10, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      shape = species #<<
      ),
    size = 2 #<<
    )
```
`shape` changes with data (mapping).

`size` is fixed for all points (setting).
]

.right-plot[
```{r, ref.label="peng10", echo=FALSE}
```
]

---

class: inverse

## Try yourself

Experiment with altering the colour, size, shape or alpha of the points in `ggplot02.R`.

---

class: middle

# Geometry layers

---

## Other layers/geometries

Instead of the `geom_point` layer, we could use something else.

Last time I counted there were 50 different geometry layers to choose from.

And there are add-on packages with more!

---

.left-code[
## `geom_point()`
```{r peng11, eval=FALSE}
ggplot(data = penguins) +
  geom_point( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

.right-plot[
```{r, ref.label="peng11", echo=FALSE}
```
]

---

.left-code[
## `geom_smooth()`
```{r peng12, eval=FALSE}
ggplot(data = penguins) +
  geom_smooth( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

.right-plot[
```{r, echo=FALSE, message=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g), alpha=0) +
  geom_smooth(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

---

.left-code[
## Adding layers
```{r peng13, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) +
  geom_smooth( #<<
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng13", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Too much typing!
```{r peng14, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    ) +
  geom_smooth(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    )
```
]

.right-plot[
```{r, ref.label="peng14", echo=FALSE, message=FALSE}
```
]

---


.left-code[
## Inheriting aesthetics
```{r peng15, eval=FALSE}
ggplot(data = penguins,
       mapping = aes( #<<
         x = flipper_length_mm, #<<
         y = body_mass_g #<<
         ) #<<
       ) +
  geom_point() +
  geom_smooth()
```
Any mapping defined in the `ggplot()` function are inherited by other layers.
]

.right-plot[
```{r, ref.label="peng15", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Unique aesthetics
```{r peng16, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         y = body_mass_g
         )
       ) +
  geom_point(
    mapping = aes( #<<
      colour = species #<<
      ) #<<
    ) +
  geom_smooth()
```
You can override or add new aesthetics in each item.
]

.right-plot[
```{r, ref.label="peng16", echo=FALSE, message=FALSE}
```
]

---

class: inverse

## Try yourself

Add some code in `ggplot02.R` to add a smoother on top of the points.

Take a look at the data visualisation cheat sheet (under the Help menu) to
find some other geometries that you might be able to use for this type of data
(two continuous measures), and try them out.

---

class: middle

# Naming things

---

## Naming things: code style

R code uses a lot of **functions**. The basic syntax for a function is:

```{r, eval=FALSE}
function_name(param1, param2, param3)
```

Functions can take a bunch of **parameters**. These are matched by name (if named), else by position.

The code for our plot contains 3 functions:
 - `ggplot(data = penguins)`
 - `geom_point(mapping = aes(...))`
 - `aes(x = flipper_length_mm, y = body_mass_g)`

Note in each case we've named the parameters (i.e. `data = penguins`, `x = flipper_length_mm`). This is generally good practice. If unnamed you're relying on order, so this works too:

```{r, eval=FALSE}
ggplot(penguins) + geom_point(aes(flipper_length_mm, body_mass_g))
```

---

## Naming things: titles and labels

Charts should, as much as possible, be self-contained. They should tell the story without the viewer/reader having to figure out too much what each bit of the plot represents.

- Concise, accurate axis labels, including units where relevant.

- Concise, clear legend names and legend items.

- A title (and possibly subtitle) that tells the story.

- Possibly further captioning that helps (e.g. data sources, authorship).

---

## Axis labels

Use `labs()` to change axis labels (or any labels associated with an aesthetic). 

Alternatively, the `xlab()` or `ylab()` functions can be used.

---

.left-code[
## Axis labels
```{r peng17, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         y = body_mass_g
         )
       ) +
  geom_point(
    mapping = aes(
      colour = species
      )
    ) +
  geom_smooth() +
  labs(x = "Flipper length (mm)", #<<
       y = "Body mass (g)") #<<
```
]

.right-plot[
```{r, ref.label="peng17", echo=FALSE, message=FALSE}
```
]

---

## Titles and captions

Titles and captions can also be changed using `labs`. This can be useful for adding data source information, or for helping the
reader interpret the chart.

You can set a chart title (and subtitles or captions) using

```{r, eval=FALSE}
labs(title = "A useful title",
     subtitle = "Shown smaller, under the title",
     caption = "Shown in the bottom-right, useful for data source")
```

---

.left-code[
## Titles tell the story
```{r peng18, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Distribution of penguin flipper lengths")
```
This is a bad title. We can tell it's a distribution already!

"Penguin" is the only useful bit.
]

.right-plot[
```{r, ref.label="peng18", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Titles tell the story
```{r peng19, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Penguin flipper length differs by species")
```
This is better.

But surely it is obvious that flipper length would differ by species!
]

.right-plot[
```{r, ref.label="peng19", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Titles tell the story
```{r peng20, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Gentoo penguins have longer flippers")
```
This is even better.

It's making the story clear.
]

.right-plot[
```{r, ref.label="peng20", echo=FALSE, message=FALSE}
```
]

---

```{r, echo=FALSE, message=FALSE, fig.dim=c(6,3.5), fig.retina=3}
border_plot <- read_csv(here::here("data/covid19/border_plot.csv"))
ggplot(border_plot) +
  aes(x=Date, y=fit) +
  geom_col(aes(x=Date, y=cases_per_arrival), fill='red') +
  geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=0.3) +
  geom_line() +
  ylab("Detected COVID-19 positives (%)") +
  xlab("") +
  theme_minimal(base_size=10) +
  ggtitle("The Mystery of the missing May: COVID-19 cases at the border",
          subtitle = "Bars are cases per person, line and band is smoothed trend with uncertainty (GAM)")
```

---

class: inverse

## Try yourself

Add titles or change axis labels as needed using `labs` in your charts from earlier.

---

class: inverse

## More exercises

Open up `quakes.R` to explore some earthquake data from around Fiji.

1. Visualise the earthquake data by longitude and latitude.

2. Try to add depth, magnitude or number of stations that the earthquake was detected at to the plot through the colour, shape, size etc aesthetics.

3. Try out some different geometries (e.g. points, density2d, bin2d). You could try altering the amount of smoothing in `geom_density2d` using the `adjust` parameter.

4. Instead of location, do a plot of magnitude versus number of stations detected. Presumably there'd be some relationship. What type of geometries might work best for that? Investigating `geom_jitter` might be useful here.
