---
title: 'Visualisation 2'
subtitle: 'Introduction to RStudio'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
penguins = na.omit(penguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

class: middle

# Scales and colours

---

## Scales

Each aesthetic has an associated scale. `x` and `y` have axes, `colour` has a guide. These scales can involve a transformation (e.g. a lot scale) or limits, or we can choose what colours to associate with each group.

We do this through the `scale_` functions. e.g.
  - `scale_x_log10()` to switch to a log scale,
  - `scale_y_continuous` to define limits on a numeric y-axis, 
  - `scale_colour_manual` for specifying colours manually.

And lots more for specific circumstances.

---

.left-code[
## Changing axis scales
```{r peng21, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) +
  scale_x_continuous( #<<
    limits = c(100, 300) #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng21", echo=FALSE, message=FALSE}
```
]

---

## More data: Horizons regional council river quality

We have counts of *E. coli* as an indicator of faecal contamination.

```{r, message=FALSE, echo=FALSE}
hrc <- read_csv(here::here("data/horizons_river_quality/horizons_river_ecoli.csv"))
hrc
```

These data are available from Land, Air, Water Aotearoa, https://www.lawa.org.nz/download-data/

---

We have *E. coli* counts from 95 sites throughout the horizons regional council.

```{r}
count(hrc, Site)
```

---

The counts can be summarised using

```{r}
summarise(hrc, min(Value), max(Value), median(Value))
```

You can see that we have very low counts (1) and very high counts (almost 200,000), with half the counts being less than 150.

These counts naturally belong on a logarithmic scale.

---

.left-code[
## Changing axis scales
```{r hrc1, eval=FALSE}
ggplot(data = hrc) +
  geom_point(
    mapping = aes(
      x = Date,
      y = Value
      ),
    alpha=0.5
  )
```
]

.right-plot[
```{r, ref.label="hrc1", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing axis scales
```{r hrc2, eval=FALSE}
ggplot(data = hrc) +
  geom_point(
    mapping = aes(
      x = Date,
      y = Value
      ),
    alpha=0.5
  ) +
  scale_y_log10() #<<
```
]

.right-plot[
```{r, ref.label="hrc2", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing axis scales
```{r hrc3, eval=FALSE}
ggplot(data = hrc,
       mapping = aes(
         x = Date,
         y = Value
       )
     ) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  scale_y_log10()
```
]

.right-plot[
```{r, ref.label="hrc3", echo=FALSE, message=FALSE}
```
]

---

class: inverse

## Try yourself

In `ggplot03.R` we're plotting the HRC data using boxplots for each of 5 sites.

Try changing the scale so it's logarithmic.

You may wish to change colours or axis labels, and add a title.

Experiment with differing geometries for this (we have a continuous variable and a discrete grouping variable). One option is
using `geom_density` as the geometry with the `x` aesthetic for counts, and either `fill` or `colour` for grouping.

---

.left-code[
## Changing colours
```{r peng_col1, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    )
```
]

.right-plot[
```{r, ref.label="peng_col1", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours
```{r peng_col2, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_manual( #<<
    values = c( #<<
      Adelie = "darkgreen", #<<
      Chinstrap = "maroon", #<<
      Gentoo = "orange" #<<
    ) #<<
  ) #<<
```
]

.right-plot[
```{r, ref.label="peng_col2", echo=FALSE, message=FALSE}
```
]

---

## Changing colours

The `c()` function is short for 'combine'. In this case we're combining 3 colours together, naming them after each species.

```{r, eval=FALSE}
c(Adelie = "darkgreen", Chinstrap = "maroon", Gentoo = "orange")
```

---

.left-code[
## Changing colours: palettes
```{r peng_col3, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_viridis_d() #<<
```
]

.right-plot[
```{r, ref.label="peng_col3", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours: palettes
```{r peng_col4, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_brewer( #<<
    palette = "Dark2" #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng_col4", echo=FALSE, message=FALSE}
```
]

---

class: inverse

## Try yourself

Add or alter some colours of the Horizons regional council plot from `ggplot03.R`.

Try switching from `SiteID` to `Site`, which has the names of the sites. You might
need to swap the `x` and `y` variables around for some plot types (e.g. box, violin or dot plot).

---

class: middle

# Qualitative data

---

## Qualitative data

You typically use `geom_bar` to summarise qualitative data (counts).

If the data are already summarised, then use `geom_col` instead ('column chart').

Order is often important. An ordered categorical variable is known as a `factor` in RStudio.

---

## Summaries of eye colour

There are two main ways categorical data comes:
- summarised into counts (one row per category)
- or individual rows per observation.

```{r}
eyes <- tibble(colour = c("blue", "grey", "green", "amber", "brown"),
               n = c(32, 15, 12, 16, 25))
eyes
```

---

## Unsummarising data

.pull-left[
The `uncount` function in `tidyr` is useful:

```{r}
long_eyes <- uncount(eyes, n)
long_eyes
```
]

.pull-right[
The `count` function in `dplyr` does the reverse:

```{r}
short_eyes <- count(long_eyes, colour)
short_eyes
```

Notice that the order has changed! We'd need to impose an order to stop this by using a `factor`
]

---

## Factors for categories

```{r}
colours <- c("blue", "grey", "green", "amber", "brown")
```

.pull-left[
```{r}
eyes <- tibble(colour = colours, #<<
               n = c(32, 15, 12, 16, 25))
eyes
eyes$colour
```
]

.pull-right[
```{r}
eyes <- tibble(colour = as_factor(colours), #<<
               n = c(32, 15, 12, 16, 25))
eyes
eyes$colour
```
]

---

.left-code[
## Bar charts in RStudio

For summarised data (i.e. counts) use `geom_col()`.

```{r bar1, eval=FALSE}
ggplot(data = eyes) +
  geom_col(mapping = aes(
    x = colour,
    y = n)
    )
```
]

.right-plot[
```{r, ref.label="bar1", echo=FALSE}
```
]

---

.left-code[
## Bar charts in RStudio

For unsummarised data use `geom_bar()`.

```{r bar2, eval=FALSE}
long_eyes <- uncount(eyes, n)
ggplot(data = long_eyes) +
  geom_bar(mapping = aes(
    x = colour
    )
  )
```

The `y` is computed by `geom_bar()`.
]

.right-plot[
```{r, ref.label="bar2", echo=FALSE}
```
]

---

.left-code[
## Bar charts in RStudio

Use `fill` to change colours.

```{r bar3, eval=FALSE}
long_eyes <- uncount(eyes, n)
ggplot(data = long_eyes) +
  geom_bar(mapping = aes(
    x = colour,
    fill = colour
    )
  )
```
]

.right-plot[
```{r, ref.label="bar3", echo=FALSE}
```
]

---

.left-code[
## Bar charts in RStudio

```{r bar4, eval=FALSE}
long_eyes <- uncount(eyes, n)
ggplot(data = long_eyes) +
  geom_bar(mapping = aes(
    x = colour,
    fill = colour
    )
  ) +
  scale_fill_manual(
    values = c(
      blue = "#1D91C0",
      grey = "#A6BDDB",
      green = "#006837",
      amber = "#885404",
      brown = "#552005"
    )
  )
```

These 'Hex codes' are red, green, blue values in hexadecimal (base 16).

Google 'colour picker' and Copy/Paste.
]

.right-plot[
```{r, ref.label="bar4", echo=FALSE}
```
]

---

class: inverse

## Try yourself

Open up `ggplot04.R` which has some summarised data on COVID-19 test performance from
a study of 4653 close contacts of cases in Guangzhou, China.

Try altering the `position` argument to `dodge`, `fill`, or `stack` for the different bar chart types.

Swapping the role of the two categorical variables (truth and test result) may help answer different questions (e.g. test performance
vs what to do after a test).

---

class: middle

# Small multiple plots

---

## Small multiple plots

We can subset the data and produce plots of each subset simultaenously with **faceting**.

The idea is that often we want to see trends or patterns between some variables, but often those trends
or patterns may vary by subgroup.

In that case we might want to plot them separately for each subgroup so we can compare and contrast.

---

## Small multiple plots

One way to do this is to filter a set of data down to subgroups and then plot each subgroup separately.

e.g. suppose we want to compare two sites on the Manawatu river for their E.coli counts over time. The river at Site80 (down by Waitoetoe park, opposite where the Turitea Stream comes out) and the river at Site15 (by the old teacher's college in Hokowhitu):

```{r}
waitoetoe = filter(hrc, Site == "manawatu at u/s pncc stp")
hokowhitu = filter(hrc, Site == "manawatu at teachers college")
```

---

## Comparing sites

.pull-left[
```{r}
summarise(waitoetoe,
          Median = median(Value),
          Max = max(Value))
```
]

.pull-right[
```{r}
summarise(hokowhitu,
          Median = median(Value),
          Max = max(Value))
```
]

In terms of the median and maximum these sites are quite similar, which makes sense - they're only about 3km from each other.

We could do a plot of each one and compare.
---

.pull-left[
```{r}
ggplot(data=waitoetoe) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma()
    )
```
]

.pull-right[
```{r}
ggplot(data=hokowhitu) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma()
    )
```
]

---

## Combining plots

The problem here is that each uses different scale. So while the patterns look kinda similar, it's hard to really
know the level as the y-axis scales are a bit different.

We could change the axes to force them to be the same using the `limits` argument to `scale_y_log10()`, but there's
quite a lot of repetition.

And, if we want them on the same scale anyway, why do we need separate y-axis scales? We could just place them side
by side and remove the additional axis space.

---

.pull-left[
```{r}
ggplot(data=waitoetoe) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma(),
    limits = c(1,20000)
    )
```
]

.pull-right[
```{r}
ggplot(data=hokowhitu) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma(),
    limits = c(1,20000)
    )
```
]

---

## Combining plots

The better way to do it is to combine the two datasets and let `ggplot2` split out the two groups for us when it creates the plot.

```{r}
two_sites =
  filter(hrc, Site %in% c("manawatu at u/s pncc stp", "manawatu at teachers college"))
```

---

.left-code[
## Combining plots

```{r facet1, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value,
                  col=Site)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  )
```

We could use colour to distinguish the sites.
]

.right-plot[
```{r, ref.label="facet1", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet2, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site)) #<<
```

But having separate plots is better.
]

.right-plot[
```{r, ref.label="facet2", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet3, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site))
```

The trend shows that the site at Waitoetoe park used to be
worse than at Hokowhitu, but now is more similar.

They're now both worse?
]

.right-plot[
```{r, ref.label="facet3", echo=FALSE}
```
]

---

## Combining plots

We can now easily generalise to more plots. e.g. what is happening with the turitea stream which flows into the Manawatu near Waitoetoe Park?

Unfortunately this was measured only since 2013, but nonetheless might be interesting

```{r}
three_sites = filter(hrc,
  Site %in% c("manawatu at u/s pncc stp",
              "manawatu at teachers college",
              "turitea at no1 dairy")
)
```

---

.left-code[
## Combining plots

```{r facet4, eval=FALSE}
ggplot(data=three_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site))
```

We can't tell if the Turitea stream is the
reason for the increased counts pre 2013 as
we don't have the measures then.

The counts post 2013 are higher in the Turitea
stream than in the river (more dilution in the river?)
]

.right-plot[
```{r, ref.label="facet4", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet5, eval=FALSE}
ggplot(data=three_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site), ncol=1) #<<
```

Can specify the number of rows or columns.

This allows us to see the correlation in the two Manawatu measures.

But is harder to assess overall levels.
]

.right-plot[
```{r, ref.label="facet5", echo=FALSE}
```
]

---

class: inverse

## Code along

Open up `ggplot5.R` which has a density plot of the body mass of penguins.

Let's alter it to split out by species or sex using `facet_wrap`.

We'll also try `facet_grid` which gives a grid of plots, where columns are one variable
and rows are another.

---

## Combining different kinds of plot

Sometimes we want to combine different types of plot. e.g. perhaps we want
to compare the body mass distribution of penguins as well as show the correlation
between bill length and depth.

We can do the separate plots and can combine them using the `patchwork` library.

(You may have to install the `patchwork` library via the Packages menu)

---

.pull-left[
```{r, warning=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5)
```
]

.pull-right[
```{r, warning=FALSE}
ggplot(penguins) +
  geom_point(aes(x=bill_length_mm,
                 y=bill_depth_mm,
                 col=species),
               alpha=0.5)
```
]

---

## Combining plots with `patchwork`

With `patchwork` we first **save** the plots into separate variables, then use `+` or `/` to combine them together.

When you save a `ggplot2` plot, you can 'print' it out by just typing the name. So your code will look like:

```{r, eval=FALSE}
g1 = ggplot(data1) + ...
g2 = ggplot(data2) + ...

g1 # To look at the first plot
g2 # To look at the second plot

library(patchwork)
g1 + g2 # To put graphs side by side
g1 / g2 # To put graphs top and bottom
```

---

.left-code[
## Combining plots with `patchwork`

```{r patch1, eval=FALSE}
library(patchwork)
g1 = ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5)

g2 = ggplot(penguins) +
  geom_point(
    aes(x=bill_length_mm,
        y=bill_depth_mm,
        col=species),
    alpha=0.5)

g1 / g2
```
]

.right-plot[
```{r, ref.label="patch1", echo=FALSE, warning=FALSE}
```
]

---

## Combining plots with `patchwork`

Use `+` to put things side by side:

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2
```

---

## Combining plots with `patchwork`

The `plot_layout()` function is useful to control things a little.

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2 + plot_layout(guides = 'collect')
```

---

The `plot_annotation()` function allows adding titles, subtitles to the total layout.

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2 + plot_layout(guides = 'collect') +
  plot_annotation(title = "Gentoo penguins are chonky and have long, narrow beaks")
```

---

Lots more can be done. See https://patchwork.data-imaginist.com/index.html for more!

```{r, echo=FALSE, warning=FALSE}
g0 = ggplot(penguins) + geom_density(aes(x=flipper_length_mm, fill=species), alpha=0.5)
```

```{r, warning=FALSE, fig.dim=c(8,3.5), fig.retina=3}
(g0 / g1 | g2) + plot_layout(guides = 'collect') +
  plot_annotation(title = "Gentoo penguins are chonky and have long, narrow beaks")
```

---

class: inverse

## Try yourself!

Load up `ggplot06.R` which has the above graph and try changing things around.

---

class: inverse

## More exercises

Open up `immunisation.R` which loads in data on immunisation coverage for each DHB by quarter since July 2009.

1. Using the `immunity8months` dataset, produce a plot of the proportion immunised at 8 months for each DHB through time, comparing them to the 95% health target.

    - The `geom_hline()` function can be used to put a horizontal line on the plot.
    - Consider how to clearly differentiate each DHB.
    - Make sure you clearly label axes and have a suitable title.

2. Choose a single DHB and filter the `immunity` data to contain all measures for that DHB in the same way the `immunity8months` one was done. Produce a plot to demonstrate how that DHB is tracking over time for all age groups.

    - Consider smoothing the data
    - Make sure you can clearly distinguish the different age groups.
    - Make sure you clearly label axes and have a suitable title.

